Subset coercions in \Coq

### 2006-05-01 papers/article/main.tex
Typing in Russel section

%\usepackage{qzapfcha}
%\usepackage{cmbright}
\usepackage{concmath}
\usepackage{ifthen}

@@

\input{typing-macros}
\input{typing-decl}
\input{typing-algo}

@@

\author{Matthieu Sozeau \\
  sozeau@lri.fr}
\title{Subset coercions in \Coq}
\date{\today}

@@

\subsubsection{The typing judgment}
\Russell{}'s typing system (figure \ref{fig:typing-algo-rules}) is based
on the \ECCfull \cite{Luo:ECC}, with sigma types but without universes.
This restriction may be removed in future work, but cause no problem for
our main purpose which is programming. The judgement $`G \typed t : T$
means $t$ is a well-typed term of type $T$ in environment $`G$, and $`G
\typed T \subd U$ means $T$ is convertible to $U$ in environment $`G$.

We need to restrict the system so that it becomes property irrelevant for
the propositions appearing in subset types. Indeed, in \PVS{} you can't
manipulate proof terms, so you have
no way to create a term of type $0 \neq 0$ from the judgment $0 :
\mysubset{x}{`N}{P}$, but in a classical Calculus of Constructions with
sigma types or inductive types, the subset type is a first class object
defined with proper introduction and
elimination terms (figure \ref{fig:subset}). What we need to do is forbid the use of these terms
in \Russell{} terms, leaving only the type former.

@@

\begin{figure*}[t]
  \typedRules
\caption{\Russell{} typing}
\end{figure*}

\subsubsection{Equivalence}
We have renamed the technique from \ps{} to 
subset \emph{equivalence} because we have a symmetric relation, contrary
to usual subtyping relations. 
because we can state reflexivity

- Not subtyping but conversion. Contravariant product.

\subsubsection{Properties}
- Assumptions, results, algo system.

>>>>

%\usepackage{qzapfcha}
%\usepackage{cmbright}
\usepackage{concmath}
\usepackage{ifthen}
\usepackage{varioref}

@@

\input{typing-macros}
\input{typing-decl}
\input{subtyping-decl}

\input{typing-algo}

@@

\author{Matthieu Sozeau \\
  LRI, Paris Sud XI University \\
  sozeau@lri.fr}
\title{Subset coercions in \Coq}
\date{\today}

@@

\subsubsection{The typing judgment}
\Russell{}'s typing system (figure \vref{fig:typing-decl-rules}) is based
on the \CCfull \cite{Barras}, with sigma types but without universes.
This restriction may be removed in future work, but cause no problem for
our main purpose which is programming. The judgement $`G \typed t : T$
means $t$ is a well-typed term of type $T$ in environment $`G$, and $`G
\typed T \subd U$ means $T$ is convertible to $U$ in environment $`G$.
Following the presentation as a pure type system \cite{PTS}, 
the set of sorts $\PTSsorts$ is defined as $\setproptype$, the
axioms are $\PTSaxioms = \{ (\Set, \Type), (\Prop, \Type) \}$ and the
rules $\PTSrules$ are defined by the following (functional) relation:
\axiomsd

We allow to form dependent sums $\Sigma x : U.V$ (\irule{Sigma}, \irule{Sum-Dep},
\irule{Pi-1}, \irule{Pi-2} rules) only when $U, V : \Set$
or $U, V : \Prop$. The first sum represents dependent pairs, useful
when defining functions returning a tuple, the latter represent
dependent pairs of propositions, which is rarely used. The dependent
pair with $U : \Set$ and $V : \Prop$ is the subset type which we
distinguish (rule \irule{Subset}). The last possible pair $U : \Prop, V : \Set$ is forbidden
because it corresponds to a pair where the last component, which is
informative, may depend on a particular proof of the proposition
$U$. This is quite contrary to the mantra ``Computations do not depend
on proofs'' which governs our programming language.

Likewise, we need to restrict the system so that it becomes property irrelevant for
the propositions appearing in subset types. Indeed, in \PVS{} you can't
manipulate proof terms, so you have
no way to create a term of type $0 \neq 0$ from the judgment $0 :
\mysubset{x}{`N}{P}$, but in a classical Calculus of Constructions with
sigma types or inductive types, the subset type is a first class object
defined with proper introduction and
elimination terms (figure \vref{fig:subset}). What we need to do is forbid the use of these terms
in \Russell{} terms, leaving only the type former.

Some basic properties of this system are well-formedness of
environments: if $`G \typed t : T$ then $\typedwf `G$ and
well-sortedness: if $`G \typed t : T$ then $T = \Type$ or $`G \typed T :
s$ for $s `: \PTSsorts$.

We now need to change the conversion rule to a stronger one which will
include our subset equivalence.

@@

\begin{figure*}[h]
  \typedRules
  \caption{\Russell{} typing}
  \label{fig:typing-decl-rules}
\end{figure*}

\subsubsection{Equivalence}
We have renamed the technique from \ps{} to 
subset \emph{equivalence} because we have a symmetric relation, contrary
to usual subtyping relations. It also permits to include $\beta$-conversion
directly in the relation. The judgment $`G \typed T \subd{} U$ (figure
\vref{fig:conversion-decl-rules}) reads $T$ is convertible to $U$ in
environment $`G$. In the following, it is always assumed that the two
input types of the conversion are well-sorted, but maybe not by the same sort.

\begin{figure}[h]
  \subtdRules
  \caption{\Russell{} conversion}
  \label{fig:conversion-decl-rules}
\end{figure}

The rule \irule{SubConv} integrates $\beta\pi$-conversion in the
judgment. It is well-known \cite{Luo, Pottinger?} that $\beta\pi$ is
confluent only on well typed terms, so we have to ensure that our rules,
when read from bottom to top, keep well-typedness. That's the purpose of
the $`G \typed U : s$ premise of the \irule{SubTrans} rule. This rules
ensures that our judgement constructs an \emph{equivalence} and has
proper modularity. The next two rules do context closure for dependent
products and sums. It is remarkable that we use contravariance for
domains in the \irule{SubProd} rule, not restricting to invariance as in
\PVS. It is accessory here, as we could have used \emph{co}variance and
still get the same judgements because we have symetry. However, it will
become important when we create coercions (see
\vref{fig:subtyping-impl-rules}).

The really interesting rules are \irule{SubSub} and
\irule{SubProof}. The first one allows to use an object of a subset type
as an object of its support type. The latter allows (maybe abusively) to
consider an object of any type as an object of any subset based on
it. By the assumption on the input types of the judgment, we now that $P$
is well-typed, but we do not care about the proof.

\subsubsection{Properties}
\begin{paragraph}{Assumptions}
  We have proved basic metatheoretical results for this system in \Coq{}
  \cite{CCP}. We have assumed that it is strongly normalizing (SN) and has the
  following subject reduction (SR) property: if $`G \typed t : T$ and $t
  "->"_{\beta\pi} t'$ then $`G \typed t' : T' \subd T$. Strong normalization
  could be proved using the generic proof method developed in
  \cite{Geuvers} but we leave it for future work. Gang Chen
  \cite{GangChen} has studied various type systems from the
  $\lambda$-cube extended by subtyping or coercive subtyping \cite{Luo},
  including the Calculus of Constructions, and proved such results as
  SN and SR for them. It requires a very careful analysis of the system 
  to avoid cycles in the proof due to the presence of dependent types and
  conversion ; hence we also leave it for future work.
\end{paragraph}

\begin{paragraph}{Results}

\end{paragraph}

\begin{paragraph}{Algorithmic typing}
\end{paragraph}
